### 2011-06-21 trunk/htdocs/references.tex
New documentation source files
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1417 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  sections.tex
%
%  Content for Ludwig documentation.
%
%  $Id$
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2008 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The Lattice Boltzmann Method}

We review here the lattice Boltzmann method applied to single
phase problems.

\subsection{The Navier Stokes Equation}

We seek to solve the isothermal Navier-Stokes equations which, often
written in vector form, express mass conservation
\begin{equation}
\partial_t \rho + \boldsymbol{\nabla}.(\rho\mathbf{u}) = 0
\label{eq_mass1}
\end{equation}
and the conservation of momentum
\begin{equation}
\partial_t (\rho\mathbf{u}) + \boldsymbol{\nabla}.(\mathbf{\rho uu}) =
-\boldsymbol{\nabla}p + \eta \nabla^2 \mathbf{u}
+\zeta \boldsymbol{\nabla}(\boldsymbol{\nabla}.\mathbf{u}).
\label{eq_momentum1}
\end{equation}
Equation~\ref{eq_mass1} expresses the local rate of change of the
density $\rho(\mathbf{r}; t)$ as the divergence of the flux of
mass associated with the velocity field $\mathbf{u}(\mathbf{r}; t)$.
Equation~\ref{eq_momentum1} expresses Newton's second law for
momentum, where the terms on the right hand side represent the
force on the fluid.

For this work, it is more convenient to rewrite these equations
in tensor notation, where Cartesian coordinates ${x,y,z}$ are
represented by indices $\alpha$ and $\beta$, viz
\begin{equation}
\partial_t \rho + \nabla_\alpha (\rho u_\alpha) = 0
\end{equation}
and
\begin{equation}
\partial_t (\rho u_\alpha) + \nabla_\beta (\rho u_\alpha u_\beta)
= -\nabla_\alpha p
+  \eta \nabla_\beta (u_\alpha \nabla_\beta + \nabla_\alpha u_\beta)
+ \zeta \nabla_\alpha (\nabla_\gamma u_\gamma).
\end{equation}
Here, repeated indices are understood to be summed over.
The conservation law is seem better if the forcing terms of the
right hand side are combined in the fluid stress $\Pi_{\alpha\beta}$
so that
\begin{equation}
\partial_t (\rho u_\alpha) +\nabla_\beta \Pi_{\alpha\beta} = 0.
\end{equation}
In this case the expanded expression for the stress tensor is
\begin{equation}
\Pi_{\alpha\beta} = p \delta_{\alpha\beta} + \rho u_\alpha u_\beta 
+ \eta \nabla_\alpha v_\beta + \zeta (\nabla_\gamma v_\gamma)\delta_{\alpha\beta}
\end{equation}
where $\delta_{\alpha\beta}$ is that of Kroneker. Th Navier-Stokes
equations have 10 degrees of freedom (or hydrodynamic modes) being
$\rho$, three components of the mass flux $\rho u_\alpha$, and 6 independent
modes from the (symmetric) stress tensor $\Pi_{\alpha\beta}$.

\subsection{The Distribution Function and its Moments}

In lattice Boltzmann, the density and velocity of the continuum fluid
are replaced by their discrete counterparrts, the distribution function
$f_i(\mathbf{r}; t)$ and the discrete velocity space $c_{i\alpha}$.
It is possible to relate the hydrodynamic modes to distribution
function via its moments, so
\begin{equation}
\rho = \sum_i f_i,  \,
\rho u_\alpha = \sum_i f_i c_{i\alpha},  \,
\Pi_{\alpha\beta} = \sum_i f_i c_{i\alpha} c_{i\beta.}
\end{equation}
here, the index of the summation is over the number of discrete
velocities used $n$, which in three dimensions is generally
denoted by D3Q$n$. We restrict our discussion to D3Q15 and
D3Q19 (see below).

In general, the number of modes supported in $n$ and these can
be written as
\begin{equation}
M^a(\mathbf{r};t) = \sum_i m_i^a f_i(\mathbf{r};t),
\end{equation}
where $m_i^a$ is the corresponding eigenvector of the collision
maxtrix (see section blah). For example, if $m_i^a = 1$ then the
mode $M^a$ is the density. Correspondingly, the distribution
function can be related to the modes $M^a$ via
\begin{equation}
f_i() = \sum_a w_i m_i^a M^a() N^a
\end{equation}

\subsubsection{Hydrodynamic modes}

\subsubsection{D3Q15 ghost modes}

\subsubsection{D3Q19 ghost modes}

\subsection{The Lattice Boltzmann Equation}

\subsection{The Collision}

\subsection{External Body Forces}

\subsection{Fluctuating LBE}

It is possible \cite{adhikari2005} to simulate fluctuating
hydrodynamics for an isothermal fluid via the inclusion of
a fluctuating stress $\sigma_{\alpha\beta}$:
\begin{equation}
\Pi_{\alpha\beta} = p\delta_{\alpha\beta} + \rho u_\alpha u_\beta
+ \eta_{\alpha\beta\gamma\delta} \nabla_\gamma u_\delta + \sigma_{\alpha\beta}.
\end{equation}
The fluctuation-dissipation theorem relates the magnitude of this
random stress to the isothermal temperature and the viscosity.

In the LBE, this translates to the addition of a random contribution
$\xi_i$ to the distribution at the collision stage, so that
\begin{equation}
\ldots + \xi_i.
\end{equation}

For the conserved modes $\xi_i = 0$. For all the non-conserved modes,
i.e., those with dissipation, the fluctuating part may be written
\begin{equation}
\xi_i (\mathbf{r}; t) = w_i m_i^a \hat{\xi}^a (\mathbf{r}; t) N^a
\end{equation}
where $\hat{\xi}^a$ is a noise termwhich has a variance determined
by the relaxation time for given mode
\begin{equation}
\left< \hat{\xi}^a \hat{\xi}^b \right> =
\frac{\tau_a + \tau_b + 1}{\tau_a \tau_b}
\left< \delta M^a \delta M^b \right>.
\label{eq_fvar}
\end{equation}

\subsubsection{Fluctuating stress}

For the stress, the random contribution to the distributions is
\begin{equation}
\xi_i = w_i \frac{Q_{i\alpha\beta} \hat{\sigma}_{\alpha\beta}}{4c_s^2}
\end{equation}
where $\hat{\sigma}_{\alpha\beta}$ is a symmtric matrix of random
variates drawn from a Gaussian distribution with variance given
by equation~\ref{eq_fvar}. In the case that the shear and bulk
viscosities are the same, i.e., there is a single relaxation
time, then the variances of the six independent components of
the matrix are given by
\begin{equation}
\left< \hat{\sigma}_{\alpha\beta} \hat{\sigma}_{\mu\nu} \right> =
\frac{2\tau + 1}{\tau^2}
(\delta_{\alpha\mu}\delta_{\beta\nu} + \delta_{\alpha\nu} \delta_{\beta\mu}).
\end{equation}

\subsubsection{Fluctuating ghost modes: D3Q15}

\subsubsection{Fluctuating ghost modes: D3Q19}

\subsection{The Propagation}

\subsection{Solid bodies and Bounce-Back on Links}

A very general method for the representation of solid objects
within the LB approach was put forward by Ladd \cite{l94a, l94b}.
Solid objects (of any shape) are defined by a boundary surface
which intersects some of the velocity vectors $\mathbf{c}_i$
joining lattice nodes. Sites inside are designated solid, while
sites outside remain fluid. The correct boundary condition is
defined by identifying \textit{links} between fluid and solid
sites, which allows those elements of the distribution which would
cross the boundary at the propagation step to be ``bounced-back''
into the fluid. This bounce-back on links is an efficient method
to obtain the  correct hydrodynamic interaction between solid
and fluid.

\subsubsection{Boundary walls}

\subsection{Colloids}

\subsubsection{General}

Colloidal particles are assumed to be spherical with a geometrical
centre $\mathbf{r}_c$, which is also the centre of mass. The
centre is allowed to move continuously across the lattice
with velocity $\mathbf{U}$; the particle has an angular velocity
$\mathbf{\Omega}$. The surface of the colloid is defined by an
input radius, $a_0$, which determines which lattice nodes are
inside or outside the colloid. (The hydrodynamic properties of
the colloid are specified by a different radius $a_h$ --- more
of this later.) The boundary links are then the set of vectors
joining lattice nodes which intersect the spherical surface
$\{\mathbf{c}_b\}$. A schematic picture is presented in
Figure~\ref{fig_coll2}. Note that a lattice node exactly at
the solid-fluid interface is defined to be outside the colloid.

In the original approach of Ladd, fluid occupied nodes both inside
and outside the particle. The effect of the ``internal fluid'' is
known to be restricted to short time scales (compared to the
characteristic time $a_0^2/\nu$), on which the fluid inside the
particle relaxes to a solid body rotation \cite{heemels}. There
are a number of problems related to the internal fluid, so we
use fully solid particles.

Nguyen and Ladd.

For each boundary link, it is useful to define a boundary vector,
$\mathbf{r}_b$, which connects the centre of the colloid to the
appropriate boundary node (note that the length of this vector is
not necessarily equal to the input radius of the colloid).

\begin{figure}[tb]
\begin{center}
\includegraphics{xfig/colloidhalflinksnew.eps}
\end{center}
\caption{Boundary links for a colloidal particle in two dimensions
with D2Q9 (half the particle is shown). Links join fluid sites (red)
to solid sites (green) and
intersect the circular shell radius $a_0$. Boundary nodes (circles)
lie exactly half way between pairs of lattice nodes. The generalisation
to three dimensions is straightforward.}
\label{fig_coll2}
\end{figure}

A \textit{boundary node} halfway between fluid nodes which are joined
by a boundary link. The position of the boundary node is always
exactly halfway along the boundary link ($\mathbf{r} + \frac{1}{2}\mathbf{c}_b
\Delta t$) regardless of the actual position of the intersection
of the colloid surface and the boundary link.
The colloid therefore has a discrete representation which becomes a
better approximation to the sphere as the input radius becomes larger;
this approximation is known to be reasonably good for $a_0 > 5\Delta x$.

\subsubsection{Bounce-back on links}

The standard boundary condition required for the solid-fluid
interface of a moving particle is described by Ladd \cite{l94b},
and is generally refered to as bounce-back on links (BBL).
A boundary link is defined as joining a node $\mathbf{r}$
inside the particle to one outside at $\mathbf{r} + \mathbf{c}_b \Delta t$.
If the post-collision distributions are denoted by $f^\ast$, then
the distributions must be reflected at the solid surface so that
\begin{equation}
\label{eq:bbl1}
f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
- \frac{2w_{c_b} \rho \mathbf{u}_b.\mathbf{c}_b}{c_s^2}
\end{equation}
where the boundary link $\mathbf{c}_{b'} = -\mathbf{c}_b$. The velocity
at the boundary
\begin{equation}
\label{eq:ub}
\mathbf{u}_b = \mathbf{U} + \mathbf{\Omega}\times\mathbf{r}_b
\end{equation}
is determined by the particle linear velocity $\mathbf{U}$ and angular
velocity $\mathbf{\Omega}$. The change in momentum described by

The force exerted on a
single link is
\begin{equation}
\mathbf{F}_b(\mathbf{r} + {\scriptstyle\frac{1}{2}}\mathbf{c}_b\Delta t;
t + {\scriptstyle\frac{1}{2}}\Delta t) = \frac{\Delta x^3}{\Delta t}
\Big[ 2f_b^\ast(\mathbf{r}; t) - \frac{2w_{c_b}\rho_0 \mathbf{u}_b .
\mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
\label{eq:fb}
\end{equation}
with corresponding torque $\mathbf{T}_b = \mathbf{r}_b \times \mathbf{F}_b$.

The total hydrodynamic force on the particle is then found by taking
the sum of
$\mathbf{F}_b$ over all the boundary links defining the particle.
The is an associated torque on each link of $\mathbf{r}_b\times\mathbf{F}_b$,
which again is summed over all links to give the total torque on the colloid.

Note that $\rho$ in the above equations is the density of the fluid
at the appropriate fluid node for the link. As the density fluctuations
in the fluid are small compared with the mean density $\rho_0$, the
density in the correction to the bounce-back can be replaced by the
mean $\rho_0$.

\subsubsection{Dynamics}
Having computed the total force and torque on an individual colloid,
is is possible to update the the linear velocities
\begin{equation}
m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t \mathbf{F} (t),
\end{equation}
where the mass of the colloid is related to the input radius by
$m_0 = {\scriptstyle\frac{4}{3}}\pi\rho_0 a_0^3$ and the angular
velocity
\begin{equation}
I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
\mathbf{T}(t),
\end{equation}
where the moment of inertia is $I_0 = {\scriptstyle\frac{2}{5}}m_o a_o^2$.
However, this explicit update is generally found to have poor stability
properties \cite{l94b, nl02}. The alternative is to use a velocity update
which is implicit \cite{heemels,nl02}.

The total force and torque on a colloid can be split into velocity-dependent
and -independent parts by combining Equations~(\ref{eq:ub})
and~(\ref{eq:fb}) to eliminate the boundary velocity $\mathbf{u}_b$.
In this way, the decomposition is
\begin{eqnarray}
\mathbf{F} = \mathbf{F}_0 - \boldsymbol{\zeta}^{FU}.\mathbf{U}
-\boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega},
\label{eq:d3}
\\
\mathbf{T} = \mathbf{T}_0 - \boldsymbol{\zeta}^{TU}.\mathbf{U}
-\boldsymbol{\zeta}^{T\Omega}.\mathbf{\Omega}.
\label{eq:d4}
\end{eqnarray}
The velocity independent parts of the force and the torque
(appropriate for a colloid at rest) are
\begin{eqnarray}
\mathbf{F}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big] \mathbf{c}_b,
\\
\mathbf{T}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big]
(\mathbf{r}_b\times \mathbf{c}_b)
\end{eqnarray}
The matrices $\boldsymbol{\zeta}$ are interpreted as drag
coefficients and can be written as
\begin{eqnarray}
\mathbf{\boldsymbol{\zeta}}^{FU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} \mathbf{c}_b \mathbf{c}_b,
\\
\mathbf{\boldsymbol{\zeta}}^{F\Omega} &=&
\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b),
\\
\mathbf{\boldsymbol{\zeta}}^{TU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b) \mathbf{c}_b,
\\
\mathbf{\boldsymbol{\zeta}}^{T\Omega} &=&
\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b)
(\mathbf{r}_b \times \mathbf{c}_b).
\end{eqnarray}
As an example, the full form of the $\boldsymbol{\zeta}^{F\Omega}$ matrix is

\begin{equation}
\label{eq:k14}
\left( \begin{array}{rrr}
\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_x &
\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_y &
\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_z \\

\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_x &
\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_y &
\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_z \\

\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_x &
\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_y &
\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_z \\

\end{array} \right).
\end{equation}
If the particle has a symmetric distribution of boundary links (the
special case where the centre of mass is on a symmetry point of the
lattice) the $\boldsymbol{\zeta}^{FU}$ and
$\boldsymbol{\zeta}^{T\Omega}$ matrices are
diagonal, while the $\boldsymbol{\zeta}^{F\Omega}$
and $\boldsymbol{\zeta}^{TU}$ matrices are zero.

The velocity updates can now be rewritten using an
implicit update, which results in
\begin{eqnarray}
\label{eq:k15}
m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t
\big[ \mathbf{F}_0 (t + {\scriptstyle\frac{1}{2}}\Delta t)
- \boldsymbol{\zeta}^{FU} . \mathbf{U}(t + \Delta t)
- \boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega}(t + \Delta t) \big],
\\
\label{eq:k16}
I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
\big[ \mathbf{T}_0(t + {\scriptstyle\frac{1}{2}} \Delta t)
- \boldsymbol{\zeta}^{TU} . \mathbf{U}(t + \Delta t)
- \boldsymbol{\zeta}^{T\Omega}. \mathbf{\Omega}(t + \Delta t) \big].
\end{eqnarray}
In general all the matrix elements are non-zero, in which case
equations~(\ref{eq:k15}) and~(\ref{eq:k16}) can be solved via a
6$\times$6 matrix inversion.

Finally, the position of the particle can be updated: an Euler
forward step is taken using the updated velocity
\begin{equation}
\mathbf{r} (t + \Delta t) = \mathbf{r} (t) + \Delta t\mathbf{U}(t + \Delta t)
\end{equation}

\subsubsection{Changes in particle shape}

One consequence is that the discrete shape of the particle fluctuates
as the colloid centre moves relative to the
lattice. This means that the size of the particle as seen by the
fluid changes despite the fact that the input radius of the colloid
is fixed.  However, these fluctuations are small for input radii
greater than about 5 lattice units \cite{l96a}. Furthermore, the
fluctuations may be greater for some input radii than others
\cite{nl02}.

In the standard approach, changes in the map of boundary links
are accomodated by the internal fluid. If the particle shape
changes so that an internal node is exposed, the internal fluid
at that site rejoins the fluid proper with the solid body momentum
(assuming the internal fluid is relaxed to solid-body rotation).
If a fluid node is covered by particle movement, then it simply
joins the internal fluid, and will relax to solid-body rotation.
Without internal fluid, changes in particle shape
in which fluid nodes are either covered or uncovered must be
accompanied by the removal or addition
of fluid with appropriate properties at the nodes in question.
It is essential that this
is done in a way which minimises the perturbation to the fluid
flow.

If a fluid node at $\mathbf{r}$ is covered by the movement of a
solid particle,
the fluid loses a mass $\rho(\mathbf{r}) \Delta x^3$, of which an excess
$\Delta M_c = (\rho(\mathbf{r}) - \rho_0)\Delta x^3$ must be replaced
explicitly so that
the overall mass density is unchanged. At the same time, the
colloid must assume the momentum lost by the fluid
$\Delta x^3 \sum_i f_i(\mathbf{r}) \mathbf{c}_i$.

Similarly, when a fluid node is exposed by particle movement,
fluid must be replaced with the appropriate distribution, density
and velocity. If the
new density is $\rho$ (to be determined by some method) then
the fluid gains an excess of mass $\Delta M_u = (\rho - \rho_0)
\Delta x^3$ which must be balanced elsewhere. If the new fluid
is assumed to have velocity $\mathbf{u}$ then the particle must
give up momentum $\Delta x^3 \rho\mathbf{u}$.

Following Nguyen and Ladd \cite{nl02}, these changes owing
to change in particle shape are implemented by added a small
correction to the bounce-back at the following time step.
The excess mass from a covered fluid site is redistributed over all the
boundary nodes by redefined the mometum transfer at bounce-back
\begin{equation}
f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
- \frac{2 w_{c_b} \rho_0 \mathbf{u}_b . \mathbf{c}_b}{c_s^2}
+ \frac{w_{c_b} \rho_0\Delta M_c }{A},
\label{eq:q1}
\end{equation}
where $A = \rho_0 \Delta x^3 \sum_b w_{c_b}$. The accompanying
force on the particle owing to the change in shape is then
that lost by the fluid plus the contributions from the final term
in (\ref{eq:q1}) summed over all the boundary links
\begin{equation}
\Delta \mathbf{F}_c = \frac{\Delta x^3}{\Delta t}\sum_{i} f_i \mathbf{c}_i
+ \frac{\Delta x^3}{\Delta t} \sum_b \frac{w_{c_b} \rho_0 \Delta M_c}{A}
\mathbf{c}_b.
\label{eq:q2}
\end{equation}

Note that for a closed surface,
$\sum_b w_{c_b}\mathbf{c}_b = 0$, and the second term on the right-hand
side of Eq.~(\ref{eq:q2}) is zero, i.e., the redistribution of mass
does not contribute to the net force and torque on the particle.
The corresponding change in torque is
\begin{equation}
\Delta \mathbf{T}_c = \frac{\Delta x^3}{\Delta t} \mathbf{r}_c
\times \sum_i f_i \mathbf{c}_i
+ \frac{\Delta x^3}{\Delta t}
\sum_b \frac{w_{c_b}\rho_0 \Delta M_c}{A} \mathbf{r}_b \times \mathbf{c}_b,
\label{eq:q2a}
\end{equation}
where $\mathbf{r}_c$ is the boundary vector at the position where the
fluid has been removed. Again, for a closed surface, the
redistribution of mass does not contribute to the net torque
on the particle.

Likewise, the contribution to the bounce-back from newly uncovered nodes
is
\begin{equation}
- \frac{w_{c_b} \rho_0 \Delta M_u}{A}
\label{eq:q3}
\end{equation}
leading to a change in force of
\begin{equation}
\Delta \mathbf{F}_u = -\frac{\Delta x^3 \rho(\mathbf{r})
\mathbf{u}(\mathbf{r})}{\Delta t} - \frac{\Delta x^3}{\Delta t} \sum_b
\frac{w_{c_b} \rho_0 \Delta M_u}{A} \mathbf{c}_b,
\label{eq:q4}
\end{equation}
with a corresponding change in the torque.
If more than one lattice node is either covered or uncovered by
the movement of the particle, then these contributions add in a simple
way. The overall contributions can be added to the
velocity-independent terms $\mathbf{F}_0$ and $\mathbf{T}_0$ appearing in
Equations~(\ref{eq:d3}) and~(\ref{eq:d4}).

\subsubsection{Particles near contact}

Particles near contact may not pocess a full set of boundary
links (see Figure~\ref{fig:f5}). This leads to a potential
non-conservation of mass associated with the particle motion
which must be corrected.

\begin{figure}[tb]
\begin{center}
\includegraphics{xfig/colloidclose.eps}
\end{center}
\caption{Two colloids close to contact are missing links in the
region of closest approach owing to a lack of fluid nodes in the
interstice.}
\label{fig:f5}
\end{figure}
Again following Nguyen and Ladd \cite{nl02}, mass conservation is
enforced explicitly using the following procedure. There is a mass
transfer associated with the bounce-back at each link of
$(2w_{c_b}\rho_0 \mathbf{u}_b . \mathbf{c}_b / c_s^2) \Delta x^3$.
The net mass transport into the particle is the sum of this
quantity over all the links, and can be written
\begin{equation}
\Delta M_s = - \frac{2\Delta x^3 \rho_0}{c_s^2}
\Big[ \mathbf{U}.\sum_b w_{c_b} \mathbf{c}_b +
  \mathbf{\Omega} . \sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b \Big].
\end{equation}
For any partilce with a full set of boundary links, both
$\sum_b w_{c_b} \mathbf{c}_b$
and $\sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b$ are zero.
However, if the set of boundary links is imcomplete
(as in Figure~\ref{fig:f5}) $\Delta M_s$ is not zero and
a correction is required. This is again made by redistributing the
excess or deficit of mass over the other existing boundary nodes.
The additional contribution to the bounce-back here is
$-w_{c_b} \rho_0 \Delta M_s/A$, where
$A = \rho_0 \Delta x^3 \sum_b w_{c_b}$ (cf.\ Equation~\ref{eq:q1}).

This contribution
now adds to the velocity-dependent part of the force and torque and
leads to a slightly different form of the drag matrices
\begin{eqnarray}
\boldsymbol{\zeta}^{FU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} (\mathbf{c}_b - \overline{\mathbf{c}_b}) \mathbf{c}_b
\\
\boldsymbol{\zeta}^{F\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b -
\overline{\mathbf{r}_b\times\mathbf{c}_b}),
\\
\boldsymbol{\zeta}^{TU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
(\mathbf{c}_b - \overline{\mathbf{c}_b}),
\\
\boldsymbol{\zeta}^{T\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
(\mathbf{r}_b\times\mathbf{c}_b - \overline{\mathbf{r}_b\times\mathbf{c}_b}).
\end{eqnarray}
The mean quantities
\begin{equation}
\overline{\mathbf{c}_b} = \frac{\sum_b w_{c_b} \mathbf{c}_b}{ \sum_b w_{c_b}}
\end{equation}
and
\begin{equation}
\overline{\mathbf{r}_b\times\mathbf{c}_b} =
\frac{\sum_b w_{c_b} \mathbf{r}_b\times\mathbf{c}_b}{\sum_b w_{c_b}}.
\end{equation}
A little algebra will show that both occurances of $\mathbf{c}_b$ and/or
$\mathbf{r}_b\times\mathbf{c}_b$ in the definition of the drag matrices
can be replaced by their deviation from the mean. This provides a
convenient way to compute the drag matrices (maintaining symmetry)
and computing the correct force and torque on the particle when close
to contact.

\subsubsection{Rotational motions}

For a number of applications it is useful to update not only
the psoition of a spherical particle, but its orientation. This
is important, e.g., when considering magnetic particles. The
best way to implementment the required rotations is by
considering quaternions.

A quaternion can be thought of as an extended complex number
\begin{equation}
q = w + xi + yj + zk
\end{equation}
where $i^2 = j^2 = k^2 = -1$ and $ijk = -1$. The norm, or length
of a quaternion is
\begin{equation}
N(q) = (w^2 + x^2 + y^2 + z^2)^{1/2}.
\end{equation}
Addition and subtraction of quaternions proceeds as normal, but
care is required for multiplication, which is associative but not
commutative. If the quaternion is writen as $q = (w, {\bf v})$ where
$\bf{v}$ is a normal 3-vector, then multiplication can be written
\begin{equation}
q_1 q_2 = (w_1 w_2 - {\bf v}_1 . {\bf v}_2, w_1 {\bf v}_2 + w_2 {\bf v}_1
+ {\bf v}_1 \times {\bf v}_2),
\end{equation}
where the usual scalar and vector products are used. Note that the
conjugate of a quaterion $q^\star = (w, -{\bf v})$ so that
$qq^\star = N^2(q)$.
It can be shown that a rotation of an arbitrary vector though an
angle $\theta$ around the unit axis of rotaion $\hat{\bf w}$ can
be represented by the quarternion
\begin{equation}
q = (\cos(\theta/2), \hat{\bf w} \sin(\theta/2))
\end{equation}
which ensures that $N(q) = 1$. If the arbitrary vector is (0, {\bf v}),
then the rotated vector ${\bf v}^{'}$ can be written as
\begin{equation}
(0, {\bf v}^{'}) = q (0, {\bf v}) q^\star
\end{equation}
which can be expanded as
\begin{equation}
{\bf v}^{'} = (1 - \cos\theta)({\bf v}.{\hat{\bf w}) \hat{\bf w} +
{\bf v}\cos\theta} + (\hat{\bf w} \times {\bf v})\sin\theta.  
\end{equation}

\subsection{Colloid-Colloid interactions}

\subsubsection{Lubrication corrections}

\subsubsection{Hard sphere interaction}

Particles do not overlap, i.e., there is always
a hard-sphere interaction which is a function of the separation $h$:
\begin{equation}
v^{hs}(h) = \left\{
\begin{array}{ll}
\infty & h \leq 0,\\
0   & h > 0.
\end{array} \right.
\end{equation}
The hard-sphere interaction does not give rise to a force.

\subsubsection{Soft-sphere interaction}

A simple soft-sphere interaction is available, the basic form
of which is:
\begin{equation}
v^{ss}(r) = \epsilon (\sigma / r)^{\nu},
\end{equation}
where $\epsilon$ sets the energy scale, and $\sigma$ sets the
characteristic width. The steepness of the potential is set by
the exponent $\nu (> 0)$.

To prevent the need for calculation of long-range interactions,
the soft-sphere potential is truncated in a ``cut-and-shift''
approach. This is done in such a way as to smoothly match
both the potential and the force at the cut-off distance
$r_c$. For $r < r_c$ the potential is then
\begin{equation}
v^{ss}(r) - v^{ss}(r_c) - (r - r_c) \left.\frac{d v^{ss}}{dr}\right|_{r=r_c}.
\label{eq_ss_shift}
\end{equation}
Clearly, this potential is not exactly what we first thought of as
soft-sphere. Matching the potential smoothly ensures conservation
of energy, while matching the force smoothly prevents potential
instabilities in the molecular dynamics update.

The soft-sphere potential can be useful as a mechanism to
keep the particles from touching (which risks hard-sphere interactions),
in which case is is possible to compute the interaction as a function
of the separation $h$, instead of $r$. The force between
two particles is computed via the derivative of equation~\ref{eq_ss_shift}
with respect to $r$:
\begin{equation}
\mathbf{F}_{ij}(r) = -\left\{\frac{d v^{ss}(r)}{dr} -
\left. \frac{d v^{ss}(r)}{dr}\right|_{r=r_c} \right\} \mathbf{\hat{r}}_{ij}
\end{equation}
where $\mathbf{\hat{r}}_{ij}$ is the unit vector joining the centre of
particle $i$ to the centre of $j$.

\subsubsection{Leonard-Jones interaction}

\section{Lattice Boltzmann for a Binary Fluid}

\subsection{The Cahn-Hilliard Equation}

\subsection{The Choice of Free Energy}

\subsection{A Second Distribution Function}

\subsection{The Collision}

\subsection{Bounce-Back on Links}

\section{Surfactants}

This section describes the approach taken to study surfactants. Here,
we always adopt the hydrid method, i.e., LB for the hydrodynmic
equations and finite difference for the associated Cahn-Hilliard
equations. Coupling between the two sectors is via the velocity
field and the external force on the fluid computed via the
divergence of the `chemical stress'.

In this model the surfactant appears as a concentration field
$\psi(\mathbf{r};t)$, alongside the compositional order parameter
$\phi(\mathbf{r};t)$ as appears in the binary fluid. There are
two corresponding Cahn-Hilliard equations: first,

\begin{equation}
\partial_t \phi + \partial_\alpha \phi u_\alpha = 
\partial_\alpha M_\phi \partial_\alpha \mu_\phi,
\end{equation}
and second
\begin{equation}
\partial_t \psi + \partial_\alpha \psi u_\alpha = 
\partial_\alpha M_\psi \partial_\alpha \mu_\psi.
\end{equation}

\subsection{Free Energy}

A number of different free energies have been used to study
surfactants [cite Yeomans]. Here, the free energy is based on
that used by van der Graff and van der Sman and others [cite].
There are a number of different contributions to the free
energy density:
\begin{equation}
f = f_\phi + f_\psi +f_1 + f_{ex}.
\end{equation}
The first is exactly that used for the binary fluid, i.e.,
\begin{equation}
f_\phi (\phi, \partial_\alpha\phi) =
{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
+ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
\end{equation}
The surface tension and the interfacial width are determined from the
choice of parameters $A,B,\kappa$ as before.
The second term in the free energy density
represents the entropic
energy of mixing of the surfactant with the bulk phases
\begin{equation}
f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
\end{equation}
where the parameter $kT$ sets the energy scale for the whole system.
The surfactant order parameter, $\psi$, is unity for a saturated
interface and zero for a plain interface.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  references.tex
%
%  Bibliography
%
%  $Id$
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2011 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
